---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
setwd("/Users/Chris/Documents/GitHub/RepData_PeerAssessment1")

# Load Packages
library(dplyr)
library(ggplot2)

# load data
activities <- read.csv("activity.csv")

# Create single Data Time Field
activities$intervalDateTime <-
    as.POSIXct(paste(activities$date, 
                     sprintf("%04d", activities$interval)),
               format("%Y-%m-%d %H %M"), 
               tz="")
```

## What is mean total number of steps taken per day?

```{r StepsPerDay}
# Get total steps per day
dailySteps <- tapply(activities$steps,activities$date,sum, na.rm=TRUE)

# Get mean and median steps overall
meanSteps <- mean(dailySteps)
medianSteps <- median(dailySteps)

# Plot histogram
hist(dailySteps, 
     main="Steps per Day", 
     xlab="Total Steps", 
     ylab="No. of Days", 
     col="grey",
     las=1)
```

The median daily steps is `r formatC(medianSteps, format="d", big.mark=",")` and the mean `r formatC(meanSteps, format="d", big.mark=",")`.

## What is the average daily activity pattern?

```{r AverageDailyActivityPattern}

# Get Average (Mean) Steps for Each Interval, regardless of Day
intervals <- aggregate(
    steps ~ interval,
    data = activities,
    FUN = mean,
    na.rm = TRUE)

#Plot Time Series
plot.ts(x = intervals$interval,
     y = intervals$steps,
     type="l", 
     xlab = "Time",
     xaxt="n",                     # Remove x axis as adding our own
     ylab = "Average Steps",
     main="Average Daily Activity Pattern")
axis(1, 
     at = seq(0, 2400, by = 100),  # Side 1 (below axis)
     las=2)                        # Labels perpendicular to axis

# Find interval with max average no of steps
maxInterval<- intervals[which.max(intervals$steps),]

```

The 5-minute interval with the maximum number of steps is `r maxInterval$interval` with an average of `r formatC(maxInterval$steps, format="d", big.mark=",")` steps.

## Imputing missing values

```{r}
# Get total rows with NAs
rowsWithNAs <- sum(is.na(activities))
```

The total number of missing values in the dataset (i.e. the total number of rows with NAs) is `r formatC(rowsWithNAs, format="d", big.mark=",")`

```{r StepsPerDayCleaned}

# Replace NAs with average for the given interval
activitiesCleaned <- activities %>%
    group_by(interval) %>%
    mutate(steps = ifelse(is.na(steps),
        mean(steps, na.rm = TRUE),
        steps))

# Get total steps per day
dailyStepsCleaned <- tapply(activitiesCleaned$steps,activitiesCleaned$date,sum, na.rm=TRUE)

# Get mean and median steps overall
meanStepsCleaned <- mean(dailyStepsCleaned)
medianStepsCleaned <- median(dailyStepsCleaned)

# Plot histogram
hist(dailyStepsCleaned, 
     main="Steps per Day (Cleaned)", 
     xlab="Total Steps", 
     ylab="No. of Days", 
     col="green",
     las=1)

```

When the data is cleaned of NAs, the median daily steps is `r formatC(medianStepsCleaned, format="d", big.mark=",")` and the mean `r formatC(meanStepsCleaned, format="d", big.mark=",")`.

This compares to a median daily steps of `r formatC(medianSteps, format="d", big.mark=",")` and mean of `r formatC(meanSteps, format="d", big.mark=",")` for the raw data.  The effect is to increase the number of steps and spread them more evenly across the days.

## Are there differences in activity patterns between weekdays and weekends?

```{r AverageDailyActivityPatternByDayType}
# Find Weekends
activitiesCleaned <- activitiesCleaned %>%
    mutate(weekday = weekdays(intervalDateTime)) %>%
    mutate(dayType = as.factor(
        ifelse(weekday == "Saturday" | weekday == "Saturday",
            "Weekend",
            "Weekday")))

# Get mean steps by Day Type and Interval
intervalsCleanedByDayType <- activitiesCleaned %>%
    group_by(dayType, interval) %>%
    summarise(meanSteps = mean(steps))

# Plot panel line chart by Weekday / Weekend                               
ggplot(intervalsCleanedByDayType,
       aes(x = interval, 
           y = meanSteps)) +
    geom_line() +
    labs(x = "Time", 
         y = "Average Steps",
         title = "Average Daily Activity Pattern") +
    facet_wrap(~ dayType, ncol = 1)

```
